CommonMind Consortium: DLPFC

### 2021-01-08 combined_analysis/combine_all_DLPFC.Rmd
add new plots

library(ggplot2)
library(dplyr)
library(ComplexHeatmap)
library(readxl)
library(gtools)
library(corrplot)
library(pvclust)
library(metafor)
library(data.table)
library(cowplot)
library(ggraph)
library(tidygraph)
library(kableExtra)
library(githubr)
library(ggrepel)
library(corrplot)
library(zenith)
library(compositions)
library(knitr)
library(patchwork)
library(plyr) 
library(BiocParallel)

@@

# Get RNASeq QCmetadata
METADATA_QC_DLPFC_ID = 'syn16816488' 
ALL_USED_IDs = METADATA_QC_DLPFC_ID
metadata = downloadFile_version(METADATA_QC_DLPFC_ID, version = 19) %>%
  dplyr::select(`Individual_ID`, Institution, Cohort, `Reported_Gender`, Sex, Ethnicity, ageOfDeath, `PMI_(in_hours)`, Dx, 
                Sample_RNA_ID, one_of('rnaSeq_isolation:RIN',
                                      'rnaSeq_dissection:Brain_Region', 
                                      'rnaSeq_isolation:RNA_Isolation_Batch',
                                      'rnaSeq_report:Ribozero_Batch', 'rnaSeq_report:Library_Batch', 
                                      'rnaSeq_report:Flowcell_Batch', 'rnaSeq_report:Exclude?',
                                      'rnaSeq_report:Mapped_Reads',
                                      'rnaSeq_report:Intragenic_Rate','rnaSeq_report:Intronic_Rate',
                                      'rnaSeq_report:Intergenic_Rate','rnaSeq_report:Genes_Detected',
                                      'rnaSeq_report:Expression_Profiling_Efficiency',
                                      'rnaSeq_report:rRNA_Rate',
                                      'rnaSeq_report:Total_Reads', 'rnaSeq_report:Percent_Aligned','rnaSeq_report:Transcripts_Detected'))

@@

### Data preprocessing
```{r preprocess.data}
# Merge metadata
METADATA = metadata %>%
  dplyr::left_join(GENOTYPE) %>%      
  dplyr::rename(Region = `rnaSeq_dissection:Brain_Region`,
                PMI = `PMI_(in_hours)`,
                RIN = `rnaSeq_isolation:RIN`, 
                ReportExclude = `rnaSeq_report:Exclude?`,
                GenotypeExclude = `SNP_report:Exclude?`,
                SampleID = Sample_RNA_ID,
                LibraryBatch = `rnaSeq_report:Library_Batch`,
                FlowcellBatch = `rnaSeq_report:Flowcell_Batch`,
                IsolationBatch = `rnaSeq_isolation:RNA_Isolation_Batch`, #isolation batch to replace ribozero batch
                RibozeroBatch = `rnaSeq_report:Ribozero_Batch`,
                MappedReads = `rnaSeq_report:Mapped_Reads`,
                IntragenicRate = `rnaSeq_report:Intragenic_Rate`, IntronicRate = `rnaSeq_report:Intronic_Rate`, 
                IntergenicRate = `rnaSeq_report:Intergenic_Rate`, GenesDetected = `rnaSeq_report:Genes_Detected`,
                ExpProfEfficiency = `rnaSeq_report:Expression_Profiling_Efficiency`, rRNARate = `rnaSeq_report:rRNA_Rate`,
                TotalReads = `rnaSeq_report:Total_Reads`, AlignmentRate = `rnaSeq_report:Percent_Aligned`, TranscriptsDetected = `rnaSeq_report:Transcripts_Detected`) %>%
  dplyr::select(SampleID, Individual_ID, Institution, Cohort, Reported_Gender,Sex, Ethnicity, ageOfDeath, PMI, Dx, Region,
                RIN, ReportExclude, GenotypeExclude, EV.1, EV.2, EV.3, EV.4, EV.5,
                LibraryBatch, FlowcellBatch, IsolationBatch, RibozeroBatch, MappedReads, TotalReads, GenesDetected, AlignmentRate,
                IntragenicRate, IntergenicRate, IntronicRate, ExpProfEfficiency, rRNARate, TranscriptsDetected) %>%
  		 		 dplyr::filter(SampleID %in% colnames(geneCountsMerged), !is.na(SampleID)) %>%
  		 		 dplyr::filter(Region %in% c('DLPFC')) %>%
			 	 dplyr::mutate(Cohort = forcats::fct_recode(Cohort, `MSSM-Penn-Pitt` = 'MSSM-Penn-Pitt', `NIMH-HBCC`='NIMH-HBCC', `NIMH-HBCC`='NIMH-HBCC-4')) %>%
			 	 dplyr::filter(Dx %in% c('Control', "SCZ"))  %>%
         dplyr::filter(ageOfDeath > 17)

@@

```{r joint.cca, fig.width=7, fig.height=7}
variables = c('Reported_Gender', 'CellType', 'ageOfDeath', 'Institution', "Dx")

@@

    fit = cancor(M1, M2)
    V_sqrt = sqrt(mean(fit$cor^2)) # Cramer's sqrt V

@@

@@

form = ~ ageOfDeath:Institution + Reported_Gender + Dx + Institution

@@

  fit = lm(Y[,id] ~ ageOfDeath:Institution + Reported_Gender + Dx + Institution, METADATA)

  data.frame( Test     = rownames(anova(fit))[1:4],
              CellType = id, 
              p        = anova(fit)[,'Pr(>F)'][1:4], 
              frac     =  calcVarPart(fit)[1:4])
})
df = data.table(do.call(rbind, df))

@@

ymax = max(df[,sum(frac),by="CellType"]$V1)*105
ggplot(df, aes(CellType, frac*100, fill=Test)) + geom_bar(stat="identity") + coord_flip() + theme_bw() + theme(aspect.ratio=1, panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +  ylab("Variance explained (%)") + theme(axis.line = element_line(colour = "transparent"), 
        axis.line.x = element_line(colour = "black"), panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), panel.border = element_blank(), 
        panel.background = element_blank(), axis.ticks.y = element_blank(), 
        legend.key = element_blank(), plot.margin = unit(c(0, 
            0.3, 0, 0.8), "cm")) + scale_y_continuous(expand=c(0,0), limits=c(0, ymax)) + xlab('')

ggplot(df, aes(CellType, Test, fill=-log10(pmax(1e-16,p)), label=ifelse(p.adj<0.05, '*', ''))) + geom_tile() + theme_bw() + scale_fill_gradient(name = bquote(-log[10]~P), low="white", high="red", limit=c(0, NA)) + theme(aspect.ratio=1, panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) + geom_text() + coord_flip()
```

```{r sex.graph, fig.width=7, fig.height=7, dev="png"}
df_cca2 = df_cca[with(df_cca, v1!=v2),]
df_cca2$key = apply(df_cca2, 1, function(x){
  x = sort(c(x['v1'], x['v2']))
  paste(x[1], x[2])
  }) 
df_cca2 = df_cca2[!duplicated(df_cca2$key),]

@@

ggraph(net,  layout = 'igraph', algorithm = 'star') + 
  geom_edge_link(aes(width = Rsq, color=Rsq), alpha = 1) +
  geom_node_point(size=20, color="grey70") +
  geom_node_text(aes(label = name), repel = FALSE) +
  theme_graph() + theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) +
  scale_edge_width(name=bquote(R^2)) +
  scale_edge_color_gradient(name=bquote(R^2), low="white", high="red", lim=c(0, max(df_cca2$V_sqrt^2))) +  ggtitle("Pairwise squared correlation")
```

```{r sex.confounding, fig.width=5, fig.height=5}
form = ~ Reported_Gender + GABA + GLU + Olig + MgAs + ageOfDeath 
par(mfrow=c(2,2))
res = lapply( unique(METADATA$Institution), function(INST){
  i = METADATA$Institution == INST
  C = canCorPairs(form, METADATA[i,]) 
  plotCorrMatrix(C, main=INST)
})
```

@@

data.frame(pi1 = 1 - qvalue(df_meta[[coef]]$P)$pi0,
          `# DE` = sum(df_meta[[coef]]$adj.P.Val < 0.05),
          check.names=FALSE ) %>% mutate_if(is.numeric, format, digits=3) %>% kable(caption="pi1 for SCZ meta-analysis") %>% kable_styling(full_width=FALSE)

@@

out_group <- inner_join(MSSM, MAYO) %>% 
  inner_join(., ROSMAP) %>% 
  inner_join(., gandal)

@@

cor <- cor(df_two)
# clustering based on distance
hcl = hclust(as.dist(1-cor), method='ward.D2')

@@

@@

```{r plot.gene_set_test.marginal, eval=TRUE}
 plot_zenith_results = function( res, coef, Resource, FDR = 0.05){

@@

  df = df[order(df$PValue, decreasing=TRUE),]
  df$Set = factor(df$Set, df$Set)
  ymax = max(-log10(df$FDR)) * 1.05
  ggplot(df, aes(Set, -log10(FDR))) + geom_bar(stat="identity", fill="navy") + theme_bw() + theme(aspect.ratio=1,plot.title = element_text(hjust = 0.5)) + coord_flip() + ylab(bquote(-log[10]~FDR)) + xlab('') + ggtitle(coef) + scale_y_continuous(expand=c(0,0), limits=c(0, ymax)) + geom_hline(yintercept=-log10(0.05), color="red", linetype="dashed")
}

@@

# Sex Analysis
```{r test.Sex, fig.height=5, fig.width=10}
# Schizophrenia testing
# extract from each chort
coef = 'Reported_GenderMale'
figList.Sex = lapply( names(df_DE_results), function(key){
  df = df_DE_results[[key]][[coef]]
  main = paste(key, coef, sep = ': ' )
  plotVolcano( df, df$Symbol[1:20] ) + ggtitle( main )
})

@@

```{r plot.meta.analysis.Sex, fig.height=5, fig.width=10}
# volcano plot for meta-analysis
main = 'Meta-analysis: Sex'
fig1 = plotVolcano( df_meta[[coef]], df_meta[[coef]]$Symbol[1:20] ) + ggtitle( main )

@@

```{r plot.gene_set_test.marginal.Sex, eval=TRUE}
 plot_zenith_results( res_meta_zenith, "Reported_GenderMale", "GO", FDR=0.05)               
 plot_zenith_results( res_meta_zenith, "Reported_GenderMale", "MSigDB", FDR=0.05)        
```

@@

```{r plot.gene_set_test.marginal.Sex.autosome, eval=TRUE}
 plot_zenith_results( res_meta_zenith.sex_autosome, "Reported_GenderMale", "GO", FDR=0.1) 
 plot_zenith_results( res_meta_zenith.sex_autosome, "Reported_GenderMale", "MSigDB", FDR=0.1)   
```

@@

```{r plot.meta.analysis.SCZ.inter_sex.disease, fig.height=5, fig.width=10}
# volcano plot for meta-analysis.  
main = paste('Meta-analysis', coef, sep = ': ' )
plotVolcano( df_meta.inter_sex.disease[[coef]], df_meta.inter_sex.disease[[coef]]$Symbol[1:20] ) + ggtitle( main )

>>>>

library(ggplot2)
library(dplyr)
library(ComplexHeatmap)
library(readxl)
library(gtools)
library(corrplot)
library(fdrtool)
library(pvclust)
library(RColorBrewer)
library(metafor)
library(data.table)
library(cowplot)
library(ggraph)
library(tidygraph)
library(kableExtra)
library(githubr)
library(ggrepel)
library(corpcor)
library(corrplot)
library(zenith)
library(compositions)
library(knitr)
library(patchwork)
library(plyr) 
library(BiocParallel)

@@

# Get RNASeq QCmetadata
METADATA_QC_DLPFC_ID = 'syn16816488' 
ALL_USED_IDs = METADATA_QC_DLPFC_ID
metadata = downloadFile_version(METADATA_QC_DLPFC_ID, version = 19) %>%
  dplyr::select(`Individual_ID`, Institution, Cohort, `Reported_Gender`, Sex, Ethnicity, ageOfDeath, `PMI_(in_hours)`, Dx, 'pH',
                Sample_RNA_ID, one_of('rnaSeq_isolation:RIN',
                                      'rnaSeq_dissection:Brain_Region', 
                                      'rnaSeq_isolation:RNA_Isolation_Batch',
                                      'rnaSeq_report:Ribozero_Batch', 'rnaSeq_report:Library_Batch', 
                                      'rnaSeq_report:Flowcell_Batch', 'rnaSeq_report:Exclude?',
                                      'rnaSeq_report:Mapped_Reads',
                                      'rnaSeq_report:Intragenic_Rate','rnaSeq_report:Intronic_Rate',
                                      'rnaSeq_report:Intergenic_Rate','rnaSeq_report:Genes_Detected',
                                      'rnaSeq_report:Expression_Profiling_Efficiency',
                                      'rnaSeq_report:rRNA_Rate',
                                      'rnaSeq_report:Total_Reads', 'rnaSeq_report:Percent_Aligned','rnaSeq_report:Transcripts_Detected'))

@@

### Data preprocessing
```{r preprocess.data}
# Merge metadata 
METADATA = metadata %>%
  dplyr::left_join(GENOTYPE) %>%      
  dplyr::rename(Region = `rnaSeq_dissection:Brain_Region`,
                PMI = `PMI_(in_hours)`,
                RIN = `rnaSeq_isolation:RIN`, 
                ReportExclude = `rnaSeq_report:Exclude?`,
                GenotypeExclude = `SNP_report:Exclude?`,
                SampleID = Sample_RNA_ID,
                LibraryBatch = `rnaSeq_report:Library_Batch`,
                FlowcellBatch = `rnaSeq_report:Flowcell_Batch`,
                IsolationBatch = `rnaSeq_isolation:RNA_Isolation_Batch`, #isolation batch to replace ribozero batch
                RibozeroBatch = `rnaSeq_report:Ribozero_Batch`,
                MappedReads = `rnaSeq_report:Mapped_Reads`,
                IntragenicRate = `rnaSeq_report:Intragenic_Rate`, IntronicRate = `rnaSeq_report:Intronic_Rate`, 
                IntergenicRate = `rnaSeq_report:Intergenic_Rate`, GenesDetected = `rnaSeq_report:Genes_Detected`,
                ExpProfEfficiency = `rnaSeq_report:Expression_Profiling_Efficiency`, rRNARate = `rnaSeq_report:rRNA_Rate`,
                TotalReads = `rnaSeq_report:Total_Reads`, AlignmentRate = `rnaSeq_report:Percent_Aligned`, TranscriptsDetected = `rnaSeq_report:Transcripts_Detected`) %>%
  dplyr::select(SampleID, Individual_ID, Institution, Cohort, Reported_Gender,Sex, Ethnicity, ageOfDeath, PMI, Dx, Region,
                RIN, ReportExclude, GenotypeExclude, EV.1, EV.2, EV.3, EV.4, EV.5, pH,
                LibraryBatch, FlowcellBatch, IsolationBatch, RibozeroBatch, MappedReads, TotalReads, GenesDetected, AlignmentRate,
                IntragenicRate, IntergenicRate, IntronicRate, ExpProfEfficiency, rRNARate, TranscriptsDetected) %>%
  		 		 dplyr::filter(SampleID %in% colnames(geneCountsMerged), !is.na(SampleID)) %>%
  		 		 dplyr::filter(Region %in% c('DLPFC')) %>%
			 	 dplyr::mutate(Cohort = forcats::fct_recode(Cohort, `MSSM-Penn-Pitt` = 'MSSM-Penn-Pitt', `NIMH-HBCC`='NIMH-HBCC', `NIMH-HBCC`='NIMH-HBCC-4')) %>%
			 	 dplyr::filter(Dx %in% c('Control', "SCZ"))  %>%
         dplyr::filter(ageOfDeath > 17)

@@

```{r summarize.plots}

METADATA %>% group_by(Dx, Institution, Reported_Gender) %>% dplyr::summarise(count = n()) %>% ggplot(aes(Institution, count, fill=Dx)) + geom_bar(stat="identity", position=position_dodge()) + facet_wrap(~ Reported_Gender) + theme_bw(16) + theme(aspect.ratio=1, axis.text.x = element_text(angle = 45, hjust=1)) + scale_fill_manual("Diagnosis", values=c("grey", "darkred"))
```

```{r summarize.plots2, fig.height=12}
variables = c( 'ageOfDeath',"RIN", "PMI", 'rRNARate', 'IntronicRate', 'IntragenicRate', 'IntergenicRate', "pH")

METADATA %>% group_by(Dx, Institution, Reported_Gender) %>% reshape2::melt(, id.vars=c('Dx', 'Institution', 'Reported_Gender'), measure.vars=variables ) %>% ggplot(aes(Institution, value, fill=Dx)) + geom_boxplot() + facet_grid( variable ~ Reported_Gender, scales='free_y') + theme_bw(16) + theme(aspect.ratio=1, axis.text.x = element_text(angle = 45, hjust=1), legend.position="none") + scale_fill_manual("Diagnosis", values=c("grey", "darkred"))
```

Wilcoxon test comparing Controls and SZ within each Institution and Sex
```{r compare.case.control, fig.height=12}
df = METADATA %>% group_by(Dx, Institution, Reported_Gender) %>% reshape2::melt(, id.vars=c('Dx', 'Institution', 'Reported_Gender'), measure.vars=variables ) %>% data.table

# test difference bween CTL and SZ within each Institution, Sex
df[,wilcox.test(value ~ Dx)$p.value, by=c("Institution", "Reported_Gender", "variable")] %>% ggplot(aes(Institution, -log10(V1))) + geom_bar(stat="identity", position=position_dodge()) + facet_grid(variable ~ Reported_Gender) + theme_bw(16) + theme(aspect.ratio=1, axis.text.x = element_text(angle = 45, hjust=1)) + scale_fill_manual("Diagnosis", values=c("grey", "darkred")) + ylab(bquote(-log[10]~P))
```

ANCOVA evaluating heterogenity in values across cohorts for each sex
```{r ancova, fig.height=12}
df2 = df[,anova(lm(value ~ Institution))$`Pr(>F)`[1], by=c("Reported_Gender", "variable")] 
df2[,p.adj := p.adjust(V1, "bonferroni")]

ggplot(df2, aes(Reported_Gender, -log10(V1), fill=p.adj < 0.05)) + geom_bar(stat="identity", position=position_dodge()) + facet_wrap(~variable, ncol=1) + theme_bw(16) + theme(aspect.ratio=1, axis.text.x = element_text(angle = 45, hjust=1)) + ylab(bquote(-log[10]~P)) + scale_fill_manual("Passes 5%\nBonferroni", values=c("grey", "navy"))

df2[,p := format(V1, digits=2,scientific=TRUE)][,c('Reported_Gender', 'variable', 'p')] %>% kable(caption="ANCOVA p-values", scientific=TRUE) %>% kable_styling(full_width=FALSE)
```

```{r joint.cca, fig.width=7, fig.height=7}

# pH is not included because it is not available om many of the samples
sum(!is.na(METADATA$pH))
sum(is.na(METADATA$pH))

variables = c('Reported_Gender', 'CellType', 'ageOfDeath', 'Institution', "Dx")
variables2 = c("RIN", "PMI", 'rRNARate', 'IntronicRate', 'IntragenicRate', 'IntergenicRate')
variables = c(variables, variables2)

@@

    # If some rows are dropped for including NA's, 
    # only analyze rows present in both variables
    rowids = intersect(rownames(M1), rownames(M2))

    fit = cancor(M1[rowids,], M2[rowids,])
    V_sqrt = sqrt(mean(fit$cor^2)) # Cramer's sqrt V

@@

<!---
Evaluate and perform hypothesis test for partial correlations
```{r pcor}
# Compute partial correlations
pC = cor2pcor(C)
colnames(pC) = colnames(C)
rownames(pC) = rownames(C)

res = lapply(colnames(C), function(v1){
  res = lapply(colnames(C), function(v2){

    r = pC[v1, v2]
    n = nrow(METADATA)
    k = ncol(C)
    t = r * sqrt((n-2-k) / (1-r^2))

    p = 2*pt( abs(t), n - 2 -k, lower.tail=FALSE)

    a = sort(c(v1, v2))

    data.frame(V1 = a[1], V2 = a[2], partial.corr = r, pvalue = p)
  })
  do.call(rbind, res)
 })
res = data.table(do.call(rbind, res))
res = res[V1 != V2,]
res[,title:=paste(V1, V2)]
res = res[,.SD[1,], by='title']

res$p.adj = p.adjust( res$pvalue, "BH")

ggplot(res, aes(V1, V2, fill=partial.corr, label=ifelse(p.adj<0.05, '*', ''))) + geom_tile() + theme_bw() + scale_fill_gradient2(name = "Partial corr", low="blue", mid="white", high="red", limit=c(-1, 1)) + theme(aspect.ratio=1, panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), legend.position="bottom") + geom_text() + coord_flip() + xlab("Variable 1") + ylab("Variable 2")
```
--->

@@

form = ~ ageOfDeath:Institution + Reported_Gender + Dx + Institution + RIN + rRNARate + PMI + IntronicRate + IntragenicRate + IntergenicRate

@@

  form2 = as.formula(paste0('Y[,id] ~ ', as.character(form)[2]))

  fit = lm(form2, METADATA)

  vp = calcVarPart(fit)
  n = length(vp) - 1 # omit residuals

  data.frame( Test     = rownames(anova(fit))[1:n],
              CellType = id, 
              p        = anova(fit)[,'Pr(>F)'][1:n], 
              frac     =  calcVarPart(fit)[1:n])
})
df = data.table(do.call(rbind, df))

@@

df2 = df[,sum(frac), by="Test"]
setorder(df2, 'V1')

df$Test = factor(df$Test, df2$Test)

ymax = max(df[,sum(frac),by="CellType"]$V1)*105
ggplot(df, aes(CellType, frac*100, fill=Test)) + geom_bar(stat="identity") + coord_flip() + theme_bw() + theme(aspect.ratio=1, panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +  ylab("Variance explained (%)") + theme(axis.line = element_line(colour = "transparent"), 
        axis.line.x = element_line(colour = "black"), panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), panel.border = element_blank(), 
        panel.background = element_blank(), axis.ticks.y = element_blank(), 
        legend.key = element_blank(), plot.margin = unit(c(0, 
            0.3, 0, 0.8), "cm")) + scale_y_continuous(expand=c(0,0), limits=c(0, ymax)) + xlab('') + scale_fill_manual(values = rev(brewer.pal( length(table(df$Test)), "Set3")))

ratio = length(CTs) / nlevels(df$Test)

ggplot(df, aes(CellType, Test, fill=-log10(pmax(1e-16,p)), label=ifelse(p.adj<0.05, '*', ''))) + geom_tile() + theme_bw() + scale_fill_gradient(name = bquote(-log[10]~P), low="white", high="red", limit=c(0, NA)) + theme(aspect.ratio=ratio, panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), legend.position="bottom") + geom_text() + coord_flip()
```

```{r sex.graph, fig.width=7, fig.height=7, dev="png"}
plotStarCorr = function(df_cca, cutoff=1e-3){

  df_cca2 = df_cca[with(df_cca, v1!=v2),]
  df_cca2$key = apply(df_cca2, 1, function(x){
    x = sort(c(x['v1'], x['v2']))
    paste(x[1], x[2])
    }) 
  df_cca2 = df_cca2[!duplicated(df_cca2$key),]

  df_cca2 = df_cca2[df_cca2$V_sqrt^2 > cutoff,]

@@

  ggraph(net,  layout = 'igraph', algorithm = 'star') + 
    geom_edge_link(aes(width = Rsq, color=Rsq), alpha = 1) +
    geom_node_point(size=20, color="grey70") +
    geom_node_text(aes(label = name), repel = FALSE) +
    theme_graph() + theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) +
    scale_edge_width(name=bquote(R^2)) +
    scale_edge_color_gradient(name=bquote(R^2), low="cornsilk", high="red", lim=c(0, max(df_cca2$V_sqrt^2))) +  ggtitle("Pairwise squared correlation")
}

plotStarCorr(df_cca, 0)
```

```{r sex.graph.sub, fig.width=7, fig.height=7, dev="png"}
i1 = with(df_cca, v1 %in% setdiff(variables, variables2))
i2 = with(df_cca, v2 %in% setdiff(variables, variables2))

plotStarCorr(df_cca[i1 & i2,], 0)
```

```{r sex.confounding, fig.width=5, fig.height=5}
form = ~ Reported_Gender + GABA + GLU + Olig + MgAs + ageOfDeath + PMI + rRNARate + RIN + IntronicRate + IntragenicRate + IntergenicRate
par(mfrow=c(2,2))
res = lapply( unique(METADATA$Institution), function(INST){
  i = METADATA$Institution == INST
  C = canCorPairs(form, METADATA[i,]) 
  plotCorrMatrix(C, main=INST)
})
```

Despite the fact that the estimated GABA fraction is lower in females, this observation is not significant after accounting for other variables, notably age of death. 
```{r sex.confounding.testing, fig.width=7, fig.height=4}

res = lapply( unique(METADATA$Institution), function(INST){
  res = lapply( c('GABA', 'GLU', 'Olig', 'MgAs'), function(CT){
    i = METADATA$Institution == INST
    y = factor(METADATA[i,]$Reported_Gender, c("Male", "Female"))

    form = as.formula(paste0('y ~ ', CT, '+ ageOfDeath + PMI + rRNARate + RIN + IntronicRate + IntragenicRate + IntergenicRate'))

    fit = glm(form,  METADATA[i,], family="binomial")

    result = data.frame(coef(summary(fit))[-1,])
    result$coef = rownames(result)
    result$Institution = INST
    result$CellType = CT
    rownames(result) = NULL

    result
    })
  do.call(rbind, res)
})
res = do.call(rbind, res)
colnames(res)[colnames(res) == 'Pr...z..'] = 'pvalue'
colnames(res)[colnames(res) == 'Std..Error'] = 'se'
res$p.adj = p.adjust(res$pvalue, "BH")

lim = max(abs(res$z.value))

res2 = res[res$coef %in%  c('GABA', 'GLU', 'Olig', 'MgAs'),]
res2$p.adj = p.adjust(res2$pvalue, "BH")

ggplot(res2, aes(Institution, Estimate, color=coef)) + geom_point() + theme_bw() + ylab("Regression coefficient") + facet_wrap(~CellType, nrow=1) + theme(aspect.ratio=1) + geom_errorbar(aes(ymin = Estimate - se, ymax = Estimate + se), width=0.1) + geom_hline(yintercept=0, linetype="dashed")
```

```{r significance.CT.diff}
colorCodes = c(
  GABA = '#66A61E',
  GLU = '#E6AB02',
  Olig = '#E7298A',
  MgAs = '#7570B3',
  Bulk = "grey50")

res2$color = colorCodes[match(res2$coef, names(colorCodes))]

ggplot(res2, aes(Institution, -log10(pvalue))) + geom_point(color=res2$color) + theme_bw() + ylab("Regression coefficient") + facet_wrap(~CellType, nrow=1) + theme(aspect.ratio=1, axis.text.x = element_text(angle = 45, hjust=1)) + geom_hline(yintercept=-log10(0.05/nrow(res2)), linetype="dashed") + ylab(bquote(-log[10]~P)) + ylim(0, 3)
```

Sow results for hypothesis testing for: `ageOfDeath + PMI + rRNARate + RIN`

```{r sex.confounding.testing2, fig.width=7, fig.height=4}

res = lapply( unique(METADATA$Institution), function(INST){
  i = METADATA$Institution == INST
  y = factor(METADATA[i,]$Reported_Gender, c("Male", "Female"))

  form = y ~ ageOfDeath + PMI + rRNARate + RIN + IntronicRate + IntragenicRate + IntergenicRate

  fit = glm(form,  METADATA[i,], family="binomial")

  result = data.frame(coef(summary(fit))[-1,])
  result$coef = rownames(result)
  result$Institution = INST
  rownames(result) = NULL

  result
})
res = do.call(rbind, res)
colnames(res)[colnames(res) == 'Pr...z..'] = 'pvalue'
res$p.adj = p.adjust(res$pvalue, "BH")

ggplot(res, aes(Institution, z.value, fill=coef)) + geom_bar(stat="identity", position = position_dodge())  + theme_bw() + ylab("Z-score from logistic regression") + theme(aspect.ratio=1)
```

Evaluating each term in the model: `Reported_Gender ~ Institution*(ageOfDeath + PMI + rRNARate + RIN + IntronicRate + IntragenicRate + IntergenicRate)` using logistic regression
```{r anova.test}
# run ANOVA LRT omitting each variable
y = factor(METADATA$Reported_Gender, c("Male", "Female"))

form = y ~ Institution*(ageOfDeath + PMI + rRNARate + RIN + IntronicRate + IntragenicRate + IntergenicRate)
fit_full = glm(form,  METADATA, family="binomial")

res = lapply(c('ageOfDeath', 'PMI', 'rRNARate','RIN', 'IntronicRate', 'IntragenicRate', 'IntergenicRate'), function(V){

  form2 = gsub(V, "1", form)
  form2 = as.formula(paste(form2[2], '~', form2[3]))

  fit_null = glm(form2,  METADATA, family="binomial")

  res = anova(fit_full, fit_null, test="Chisq")

  data.frame(Variable = V, pvalue = res$`Pr(>Chi)`[2])
})
res = do.call(rbind, res)
res$p.adj = p.adjust(res$pvalue, "bonferroni")

res %>% kable(caption="Hypothesis test ", digits=8) %>% kable_styling(full_width=FALSE)
```

@@

data.frame(pi1 = 1 - qvalue(df_meta[[coef]]$P)$pi0,
          `# DE` = sum(df_meta[[coef]]$adj.P.Val < 0.05),
          check.names=FALSE ) %>% mutate_if(is.numeric, format, digits=3) %>% kable(caption="pi1 for SCZ meta-analysis") %>% kable_styling(full_width=FALSE)
```

@@

# out_group <- inner_join(MSSM, MAYO) %>% 
#   inner_join(., ROSMAP) %>% 
#   inner_join(., gandal)

out_group <- inner_join(gandal, ROSMAP)

@@

cor <- cor(df_two, method="spearman")
# clustering based on distance
hcl = hclust(as.dist(1-cor), method='ward.D2')

@@

## t-statistics for CASKIN1
```{r CASKIN1}
 df_two['ENSG00000167971',]  %>% kable(caption="t-statistics for CASKIN1", row.names=FALSE) %>% kable_styling(full_width=FALSE)
```

@@

```{r plot.gene_set_test.marginal, eval=TRUE}

  plot_zenith_results = function( res, coef, Resource, FDR = 0.05, compare=NULL){

@@

  # filter by FDR
  df$FDR = pmax(1e-300, df$FDR)
  df = df[df$FDR <= FDR,]
  df$Coef = c()
  df$se = c()
  df$Resource = c()

  df = df[order(df$PValue, decreasing=TRUE),]
  df$Set = factor(df$Set, df$Set)
  ymax = max(-log10(df$FDR)) * 1.05

  if( !is.null(compare) ){
    df$Direction = compare[(sign(df$effect) + 1)/2 + 1]
    fig = ggplot(df, aes(Set, -log10(FDR), fill = Direction )) + geom_bar(stat="identity") + scale_fill_manual(name = "Upregulated in", values = c('red', 'blue'))
  }else{
    fig = ggplot(df, aes(Set, -log10(FDR) )) + geom_bar(stat="identity", fill="navy")
  }

  fig + theme_bw() + theme(aspect.ratio=1,plot.title = element_text(hjust = 0.5)) + coord_flip() + ylab(bquote(-log[10]~FDR)) + xlab('') + ggtitle(coef) + scale_y_continuous(expand=c(0,0), limits=c(0, ymax)) + geom_hline(yintercept=-log10(0.05), color="black", linetype="dashed") 
}

@@

# Sex Analysis
```{r test.Sex, fig.height=5, fig.width=10}
# Schizophrenia testing
# extract from each chort
coef = 'Reported_GenderMale'
figList.Sex = lapply( names(df_DE_results), function(key){
  df = df_DE_results[[key]][[coef]]
  main = paste(key, coef, sep = ': ' )
  plotVolcanoSex( df, df$Symbol[1:20] ) + ggtitle( main )
})

@@

```{r plot.meta.analysis.Sex, fig.height=5, fig.width=10}
# volcano plot for meta-analysis
main = 'Meta-analysis: Sex'
fig1 = plotVolcanoSex( df_meta[[coef]], df_meta[[coef]]$Symbol[1:15] ) + ggtitle( main )

@@

Evaluate if GABA pathways are different between sexes
```{r zenith.gaba}
df = data.table(res_meta_zenith)
df = df[Coef == 'Reported_GenderMale',] 
df2 = df[grep("GABA", Set),]

# correct only for number of tests that have GABA in the name
df2[,FDR := p.adjust(PValue, "fdr")] 
df2 %>% kable(caption="Test only gene sets with GABA in the name") %>% kable_styling(full_width=FALSE)
```

```{r plot.gene_set_test.marginal.Sex, eval=TRUE, fig.width=12}
  plot_zenith_results( res_meta_zenith, "Reported_GenderMale", "GO", FDR=0.05, compare=c('Female', "Male"))               
  plot_zenith_results( res_meta_zenith, "Reported_GenderMale", "MSigDB", FDR=0.05, compare=c('Female', "Male"))         
```

Show genes in neurexin gene set
```{r neurexin}
geneNames = gs.GO.collection[[grep('GO0042043', names(gs.GO.collection))]]

df = df_meta[['Reported_GenderMale']]

i = match( geneNames, trim_ensembl_ids(rownames(df)))

df = df[i[!is.na(i)],]
df$ischrY = df$Chrom == "Y"

figList = lapply( unique(df$ischrY), function(TF){

  df2 = df[df$ischrY == TF,]
  ggplot(df2, aes(Symbol, logFC, color=adj.P.Val < 0.05)) + geom_point() + geom_errorbar(aes(ymin = logFC - se, ymax = logFC + se, width=.1)) + theme_bw(16) + theme(aspect.ratio=1) + geom_hline(yintercept=0, linetype="dashed") + coord_flip() + scale_color_manual(name="FDR < 5%", values=c("black", "red"))
  })
plot_grid(plotlist = figList, ncol=1)
```

@@

```{r plot.gene_set_test.marginal.Sex.autosome, eval=TRUE}
  plot_zenith_results( res_meta_zenith.sex_autosome, "Reported_GenderMale", "GO", FDR=0.1, compare=c('Female', "Male"))  
  plot_zenith_results( res_meta_zenith.sex_autosome, "Reported_GenderMale", "MSigDB", FDR=0.1, compare=c('Female', "Male"))    
```

@@

```{r plot.meta.analysis.SCZ.inter_sex.disease, fig.height=5, fig.width=10}
# volcano plot for meta-analysis.  
main = paste('Meta-analysis', coef, sep = ': ' )
plotVolcano( df_meta.inter_sex.disease[[coef]], df_meta.inter_sex.disease[[coef]]$Symbol[1:20] ) + ggtitle( main )  
```

Here we used multiple methods (BH, qvalue and FDRTool), to compute the local and global FDR and evalute the number of genes passing the 25% cutoff.
```{r pv.hist.fdrtool}
pv = df_meta.inter_sex.disease[[coef]]$P

res_fdr = fdrtool( pv, "pvalue", plot=FALSE, verbose=FALSE)

cutoff = .25
df = data.frame(`BH` = sum(p.adjust(pv, "BH") < cutoff),
          `qvalue (global)` = sum(qvalue(pv)$qvalue < cutoff),
          `qvalue (local)` = sum(qvalue(pv)$lfdr < cutoff),
          `FDRtool (global)` = sum(res_fdr$qval < cutoff),
          `FDRtool (local)` = sum(res_fdr$lfdr < cutoff),
          check.names=FALSE )  %>% t 
colnames(df) = "Count DE genes"

df %>% kable(caption="Number of DE genes by each multiple testing method at cutoff of 25%") %>% kable_styling(full_width=FALSE)
```

The pi1 value is `r 1 - qvalue(pv)$pi0`.
